Disturbance-aware minimum-time planning strategies for motorsport vehicles with probabilistic safety certificates

### 2025-05-21 performance_comparison.tex
[MM] constraints name changed and some results description

\begin{figure}
	\centering
	\includegraphics{Fig/track.eps}
	\caption{Sector of the Catalunya circuit considered in the analysis, corresponding to the curvilinear abscissa interval $\left[0.70, 0.77\right]$. Checkpoints are indicated by labels and are uniformly spaced by $\Delta\al=0.01$. This sector includes two distinct corners: a low-speed turn from $\al = 0.72$ to $\al = 0.73$, and a high-speed turn from $\al = 0.75$ to $\al = 0.76$, allowing for the evaluation of vehicle behavior across different dynamic scenarios.}
	\label{fig:track}
\end{figure}

\subsection{Robustified constraints comparison}
It is of interest to analyze how a robustified constraint---or a combination of the two described in Sections~\ref{sec:adherence_constraint} and~\ref{sec:stayontrack_constraint}---influences the driving style. This analysis compares the nominal trajectory, obtained without robustified constraints, with those resulting from a robustified stay-on-track constraint (denoted as SoT), a robustified adherence limit constraint (denoted as Adh), and a scenario where both constraints are robustified (denoted as SoT+Adh). All the optimization are obtained with the open-loop method described in Section~\ref{sec:open_loop_planning}, setting $H=4$, and $\ga^\textrm{SoT}=\ga^\textrm{Adh}=1.28$, corresponding to a 90\% probability of meeting both constraints.

Figure~\ref{fig:ol_telemetries} presents the variation of four signals with respect to the nominal trajectory, for the three previously mentioned cases. The panels display the longitudinal speed $u$ (first panel), wheel steering angle $\delta$ (second panel), total longitudinal force $F_x$ (third panel), and lateral deviation of the center of mass from the track centerline $e$ (fourth panel). Except for the $u$ panel, each plot includes dashed black lines that indicate the sign of the nominal signal-whether it is positive, negative, or numerically zero. These sign indicators are represented as three distinct levels---high, zero, and low---corresponding to positive, zero, and negative values of the nominal signal, respectively. The levels are scaled appropriately in each plot to ensure clarity of visualization.

\begin{figure}
	\centering
	\includegraphics{Fig/ol_telemetries.eps}
	\caption{}
	\label{fig:ol_telemetries}
\end{figure}

From the first panel, it can be observed that all configurations maintain a lower mean value of the longitudinal speed compared to the nominal trajectory, resulting in an increased sector time. The two configurations incorporating the robustified stay-on-track constraint exhibit, as expected, a significantly altered CoM trajectory. This behavior explains the higher longitudinal speed observed around $\alpha = 0.71$, which is followed by a sharp reduction.

@@

>>>>

\begin{figure}
	\centering
	\includegraphics{Fig/track.pdf}
	\caption{Sector of the Catalunya circuit considered in the analysis, corresponding to the curvilinear abscissa interval $\left[0.70, 0.77\right]$. Checkpoints are indicated by labels and are uniformly spaced by $\De\al=0.01$. This sector includes two distinct corners: a low-speed turn from $\al = 0.72$ to $\al = 0.73$, and a high-speed turn from $\al = 0.75$ to $\al = 0.76$, allowing for the evaluation of vehicle behavior across different dynamic scenarios.}
	\label{fig:track}
\end{figure}

\subsection{Open loop parameters sensitivity}
\label{sec:ol_param_sensitivity}
In this section, we explore the open-loop approach to gain deeper insight into the influence of key parameters on the results. For this analysis, we consider optimizations in which only the stay-on-track constraint is robustified. This choice is motivated by the fact that the effects of the stay-on-track constraint are more visually evident compared to those of the adherence limit constraint, thereby facilitating a more immediate understanding of their impact.

The first parameter analyzed is the factor $\ga^\textrm{OTC}$, which multiplies the standard deviation of the constraint, $\sigma^\textrm{OTC}$, to determine the total back-off term $\be^\textrm{OTC}$. This parameter directly influences the probability of satisfying the stay-on-track constraint: higher values of $\ga^\textrm{OTC}$ correspond to more conservative (i.e., robust) behavior. Specifically, $\ga^\textrm{OTC} = 0$ yields a satisfaction probability of 50\%, while $\ga^\textrm{OTC} = 3$ corresponds to 99\%. For the purposes of this analysis, we examine the effect of varying $\ga^\textrm{OTC}$ within this range.

The second parameter examined is the initial value of the covariance matrix, $\bP_0$. In addition to the baseline value $\bar{\bP}_0$, two alternative values are considered: $\bP_0 = 4\bar{\bP}_0$ and $\bP_0 = \frac{1}{4}\bar{\bP}_0$. These correspond, respectively, to doubling and halving the initial uncertainty associated with the state vector.

\begin{figure}
	\centering
	\includegraphics{Fig/gamma_sensitivity.pdf}
	\hfill
	\includegraphics{Fig/Pzero_sensitivity.pdf}
	\caption{}
	\label{fig:ol_sensitivities}
\end{figure}

\subsection{Robustified constraints comparison}
It is of interest to analyze how a robustified constraint---or a combination of the two described in Sections~\ref{sec:adherence_constraint} and~\ref{sec:ontrack_constraint}---influences the driving style. This analysis compares the nominal trajectory, obtained without robustified constraints, with those resulting from a robustified stay-on-track constraint (denoted as OTC), a robustified adherence limit constraint (denoted as FLC), and a scenario where both constraints are robustified (denoted as OTC+FLC). All the optimization are obtained with the open-loop method described in Section~\ref{sec:open_loop_planning}, setting $H=4$, and $\ga^\textrm{OTC}=\ga^\textrm{FLC}=1.28$, corresponding to a 90\% probability of meeting both constraints.

Figure~\ref{fig:ol_telemetries} presents the variation of four signals with respect to the nominal trajectory, for the three previously mentioned cases. The panels display the longitudinal speed $u$ (first panel), wheel steering angle $\de$ (second panel), total longitudinal force $F_x$ (third panel), and lateral deviation of the center of mass from the track centerline $e$ (fourth panel). Except for the $u$ panel, each plot includes dashed black lines that indicate the sign of the nominal signal-whether it is positive, negative, or numerically zero. These sign indicators are represented as three distinct levels---high, zero, and low---corresponding to positive, zero, and negative values of the nominal signal, respectively. The levels are scaled appropriately in each plot to ensure clarity of visualization.

\begin{figure}
	\centering
	\includegraphics{Fig/ol_telemetries.pdf}
	\caption{}
	\label{fig:ol_telemetries}
\end{figure}

From the first panel, it can be observed that all configurations maintain a lower mean value of the longitudinal speed compared to the nominal trajectory, resulting in an increased sector time. The two configurations incorporating the robustified stay-on-track constraint exhibit, as expected, a significantly altered CoM trajectory. This behavior explains the higher longitudinal speed observed around $\al = 0.71$, which is followed by a sharp reduction.

@@

It can be interesting to notice that, in the third panel, only the configuration with both the constraints robustified needs to reduce the longitudinal force during the high speed turn, in the curvilinear abscissa interval $\al\in\left[0.75, 0.76\right]$, to meet both the constraints. 
The vehicle constrained by the stay-on-track condition is forced to follow a wider trajectory, which entails greater lateral acceleration and, consequently, a higher lateral force demand---resulting in increased tire utilization. Due to the limitations imposed by the robust adherence constraint, it is necessary to reduce the available accelerating force.

\begin{figure}
	\centering
	\includegraphics{Fig/ol_saturation.pdf}
	\caption{}
	\label{fig:ol_saturation}
\end{figure}

\subsection{Simulations with noise realization}
This analysis aims to provide empirical validation of the robust trajectories obtained using the methods proposed in this work. For conciseness, the comparison is limited to a nominal trajectory and a trajectory computed using the closed-loop method with only the on-track constraint robustified.
An LQR controller, designed according to the procedure outlined in Section~\ref{sec:LQR}, is implemented along the nominal trajectory. In contrast, the robust trajectory is tracked using the controller directly obtained from the optimization process.

\begin{figure}
	\centering
	\includegraphics{Fig/olcl_traj_strings.pdf}
	\caption{}
	\label{fig:traj_strings}
\end{figure}

### 2025-05-21 stochastic_vehicle_dynamics.tex
[MGa] paper's contributions and organization
[MM] constraints name changed and some results description

\section{TMP NAME}
\label{sec:tmp}

@@

\subsection{Robust adherence limit constraint formulation}
\label{sec:adherence_constraint}
The model employs a tire formulation that neglects combined slip effects, i.e., it does not account for the simultaneous utilization of longitudinal and lateral tyre forces. As a result, the optimal control problem must include an additional constraint to ensure that the total ground reaction forces remain within the bounds of the tyre's adherence ellipse.

The base version of the constraint for each axle can be expressed as a function of $\bx$ and $\bu$, as follows:
\begin{equation}
	h^\textrm{A}_j(\bx,\bu) = \left[ \left( \frac{X_j(\bx,\bu)}{\mu_{x,j}} \right)^2+ \left( \frac{Y_j(\bx,\bu)}{\mu_{y,j}}\right)^2\right]\frac{1}{Z_j^2(\bx,\bu)}-1\leq0\qquad j=1,2.\label{eq:adherence_constraint}
\end{equation}
where the superscript A denotes adherence, and the subscript $j$ indicates the axle index, with $j=1$ referring to the front axle and $j=2$ to the rear axle. Here, $X_j$ and $Y_j$ denote the longitudinal and lateral components of the in-plane ground forces, respectively, as illustrated in Figure~\ref{fig:vehicle_model}, while $Z_j$ represents the total vertical load acting on the axle. 
As can be seen from~\eqref{eq:adherence_constraint}, the points $\left(X_j,Y_j\right)$ are constrained to lie within an ellipse whose semi-axes are given by $\mu_{x,j}Z_j$ and $\mu_{y,j}Z_j$. The constraint, without the back-off term, is represented in the left panel of Figure~\ref{fig:robust_constraints} by the solid line ellipse in the plane $X_j-Y_j$.

To better clarify the effect of the back-off term on the constraint we can suppose to have a combination of $\bx$, $\bu$, $P$ and $\ga^\textrm{A}_j$ so that $\be^\textrm{A}_j=0.2$. This implies that the semi-axes of the ellipse are reduced by a factor $\sqrt{1-\be^\textrm{A}_j} = \sqrt{0.8}$, as shown in the left panel Figure~\ref{fig:robust_constraints}, where the ellipse in dashed line delimits the available region with the back-off applied. This allows the optimizer to determine the most appropriate trade-off between the longitudinal force $X_j$ and the lateral force $Y_j$ --- often by reducing both components to some extent --- based on the specific requirements of the manoeuvre.

@@

\begin{figure}
	\centering
	\input{Fig/Tikz_Sources/tikz_adherence_constraint.tex}
	\hfill
	\input{Fig/Tikz_Sources/tikz_stayontrack_constraint.tex}
	\caption{Graphical representation of the two constraints analyzed in the section. The left panel shows the adherence constraint in the plane $X_j-Y_j$. The ellipse in solid line, whose semi-axes are given by $\mu_{x,j}Z_j$ and $\mu_{y,j}Z_j$, represents the border of the available region for the base constraint, without back-off. The ellipse in dashed line delimits the available region when a back-off $\be^\textrm{A}_j=0.2$ is present. This ellipse has its semi-axes reduced by a factor $\sqrt{1-\be^\textrm{A}_j}$ w.r.t. the one in solid line. 
	}
	\label{fig:robust_constraints}
\end{figure}

\subsection{Stay-on-track constraint}
\label{sec:stayontrack_constraint}
A stay-on-track constraint is required to ensure that the vehicle remains within the circuit boundaries. The formulation adopted introduces an algebraic variable $e$, which satisfies the following constraint:
\begin{equation}
	\begin{Bmatrix}
		x\\y
	\end{Bmatrix}
	-\boldsymbol{c} - e\bn=\boldsymbol{0} \label{eq:onplane_constraint}
\end{equation}
%\begin{equation}
%	\bp-\boldsymbol{c}_k - e_k\bn_k=\boldsymbol{0} \label{eq:onplane_constraint}
%\end{equation}
where $\boldsymbol{c}$ indicates the position of the centerline, and $\bn$ is the unit normal vector to the centerline. Equation~\eqref{eq:onplane_constraint} constraints the CoM of the vehicle to lie on the vertical plane defined by the unit normal vector $\bn$. The previously introduced algebraic variable $e$ is further constrained to take values between $e_\textrm{min}$ and $e_\textrm{max}$, which are determined based on the width of the circuit and the vehicle's front and rear tracks.

In order to take into account for the back-off terms, the maximum and minimum values are modified coherently, so that:
\begin{equation}
	e_\textrm{min} + \be^\textrm{SOT} \leq e \leq e_\textrm{max} - \be^\textrm{SOT}\label{eq:stayontrack_constraint}
\end{equation}
where $\be^\textrm{SOT}$ is computed by considering that the gradient of the constraint w.r.t. the state vector $\na_{\bx}h^\textrm{SOT}$ depends solely on $\bn$.
%Adding a back-off to this constraint produces the same effect as an increase of the front and rear tracks of the vehicle does.

>>>>

\section{Stochastic vehicle dynamic framework}
\label{sec:svdf}

@@

\subsection{Robust friction limit constraint formulation}
\label{sec:adherence_constraint}
The model employs a tire formulation that neglects combined slip effects, i.e., it does not account for the simultaneous utilization of longitudinal and lateral tyre forces. As a result, the optimal control problem must include an additional constraint to ensure that the total ground reaction forces remain within the bounds of the tyre's adherence ellipse.

The base version of the constraint for each axle can be expressed as a function of $\bx$ and $\bu$, as follows:
\begin{equation}
	h^\textrm{FLC}_j(\bx,\bu) = \left[ \left( \frac{X_j(\bx,\bu)}{\mu_{x,j}} \right)^2+ \left( \frac{Y_j(\bx,\bu)}{\mu_{y,j}}\right)^2\right]\frac{1}{Z_j^2(\bx,\bu)}-1\leq0\qquad j=1,2.\label{eq:adherence_constraint}
\end{equation}
where the superscript FLC denotes friction limit constraint, and the subscript $j$ indicates the axle index, with $j=1$ referring to the front axle and $j=2$ to the rear axle. Here, $X_j$ and $Y_j$ denote the longitudinal and lateral components of the in-plane ground forces, respectively, as illustrated in Figure~\ref{fig:vehicle_model}, while $Z_j$ represents the total vertical load acting on the axle.
As can be seen from~\eqref{eq:adherence_constraint}, the points $\left(X_j,Y_j\right)$ are constrained to lie within an ellipse whose semi-axes are given by $\mu_{x,j}Z_j$ and $\mu_{y,j}Z_j$. The constraint, without the back-off term, is represented in the left panel of Figure~\ref{fig:robust_constraints} by the solid line ellipse in the plane $X_j-Y_j$.

To better clarify the effect of the back-off term on the constraint we can suppose to have a combination of $\bx$, $\bu$, $P$ and $\ga^\textrm{FLC}_j$ so that $\be^\textrm{FLC}_j=0.2$. This implies that the semi-axes of the ellipse are reduced by a factor $\sqrt{1-\be^\textrm{FLC}_j} = \sqrt{0.8}$, as shown in the left panel Figure~\ref{fig:robust_constraints}, where the ellipse in dashed line delimits the available region with the back-off applied. This allows the optimizer to determine the most appropriate trade-off between the longitudinal force $X_j$ and the lateral force $Y_j$ --- often by reducing both components to some extent --- based on the specific requirements of the manoeuvre.

@@

\begin{figure}
	\centering
	\input{Fig/Tikz_Sources/tikz_adherence_constraint.tex}
	\hfill
	\input{Fig/Tikz_Sources/tikz_stayontrack_constraint.tex}
	\caption{Graphical representation of the two constraints analyzed in the section. The left panel shows the friction limit constraint in the plane $X_j-Y_j$. The ellipse in solid line, whose semi-axes are given by $\mu_{x,j}Z_j$ and $\mu_{y,j}Z_j$, represents the border of the available region for the base constraint, without back-off. The ellipse in dashed line delimits the available region when a back-off $\be^\textrm{FLC}_j=0.2$ is present. This ellipse has its semi-axes reduced by a factor $\sqrt{1-\be^\textrm{FLC}_j}$ w.r.t. the one in solid line.
	}
	\label{fig:robust_constraints}
\end{figure}

\subsection{On-track constraint}
\label{sec:ontrack_constraint}
An on-track constraint is required to ensure that the vehicle remains within the circuit boundaries. The formulation adopted introduces an algebraic variable $e$, which satisfies the following constraint:
\begin{equation}
	\begin{Bmatrix}
		x\\y
	\end{Bmatrix}
	-\boldsymbol{c} - e\bn=\boldsymbol{0} \label{eq:onplane_constraint}
\end{equation}
%\begin{equation}
%	\bp-\boldsymbol{c}_k - e_k\bn_k=\boldsymbol{0} \label{eq:onplane_constraint}
%\end{equation}
where $\boldsymbol{c}$ indicates the position of the centerline, and $\bn$ is the unit normal vector to the centerline. Equation~\eqref{eq:onplane_constraint} constraints the CoM of the vehicle to lie on the vertical plane defined by the unit normal vector $\bn$. The previously introduced algebraic variable $e$ is further constrained to take values between $e_\textrm{min}$ and $e_\textrm{max}$, which are determined based on the width of the circuit and the vehicle's front and rear tracks.

In order to take into account for the back-off terms, the maximum and minimum values are modified coherently, so that:
\begin{equation}
	e_\textrm{min} + \be^\textrm{OTC} \leq e \leq e_\textrm{max} - \be^\textrm{OTC}\label{eq:ontrack_constraint}
\end{equation}
where $\be^\textrm{OTC}$ is computed by considering that the gradient of the constraint w.r.t. the state vector $\na_{\bx}h^\textrm{OTC}$ depends solely on $\bn$.
%Adding a back-off to this constraint produces the same effect as an increase of the front and rear tracks of the vehicle does.
